---
title: "Introduction to prioritization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to prioritization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `prioritization` package provides spatial and non-spatial tools to assist with spatial planning and investment prioritization workflows.

In this vignette, we will walk through the two core multi-criteria evaluation functions currently in the package: configuring your constraints (`s_config`) and standardizing your inputs (`s_standardize`).

## Setup

First, load the package:

```{r setup}
library(prioritization)
```

## Standardizing Data

Continuous values often need to be standardized to a 0-100 scale before they can be weighted and combined. We can do this using `s_standardize()`.

The function supports two scaling methods: `"observe"` (scaling purely based on the local minimum and maximum of the dataset) and `"benchmark"` (scaling based on absolute reference values that you provide).

```{r}
# A simple vector of costs
costs <- c(150, 200, 350, 500)

# Observe method (relative scaling)
s_standardize(costs, selected_method = "observe")

# Benchmark method (absolute scaling against limits of 0 and 1000)
s_standardize(costs, selected_method = "benchmark", 
              min_bound = 0, max_bound =1000)
```

## Configuring Parameters for Prioritization

Before assembling a spatial score, you need a configuration `data.frame` mapping variable behaviors (direction, weights, standardization bounds). You can generate this mapping directly from your raw dataset using `s_config()`.

```{r}
# Suppose we have dataset with population and infrastructure cost
planning_data <- data.frame(
  pop_density = c(10, 50, 100, 20),
  cost = c(500, 1500, 800, 1000)
)

# Tell the config that higher population is better, but lower cost is better
my_config <- s_config(
  planning_data,
  direction = c("larger is better", "smaller is better")
)

my_config
```

The resulting configuration table can then be used in downstream spatial prioritization scripts to ensure all spatial layers are standardized and aggregated correctly.

